---
title: "Lab 12 Activity"
format: pdf
---


```{r echo = FALSE, eval = FALSE}
## the following code should be run first (from the main lab)
library(tidyverse)
library(haven)
library(caret)
library(psych)
library(paran)
library(lavaan)
dat <- read_sav("Lab_12_Cox.sav")
```


The dataset that includes the Panas items also includes the Relationship Scales Questionnaire (RSQ), which includes 30 items on 5-point Likert scales (scored 1-5). Although the entire set of 30 items can be thought of as a single scale, the following subscales are also theoretically present:

* RSQ items 11, 18, 21, 23, 25 : attachment avoidance
* RSQ items 10, 12, 13, 15, 20, 24, 29, 30: attachment anxiety

Here, RSQ items 12, 13, 20, 24, 29 need to be reverse coded, which can be done by running the following code:

```{r eval = FALSE}
dat <- dat %>% mutate(RSQ12 = 6 - RSQ12, RSQ13 = 6 - RSQ13, 
                      RSQ20 = 6 - RSQ20, RSQ24 = 6 - RSQ24, 
                      RSQ29 = 6 - RSQ29)
dat_for_fa <- dat %>% 
  dplyr::select(RSQ11, RSQ18, RSQ21, RSQ23, RSQ25,
                RSQ10, RSQ12, RSQ13, RSQ15, RSQ20, 
                RSQ24, RSQ29, RSQ30)
```

1. First, split the data into a training (exploratory) set and a test (confirmatory) set. Use a different split than used in the lab slides by setting a different seed.

```{r echo = FALSE, eval = FALSE}
set.seed(255) # answer will vary for this and all results due to different seeds
idx <- createFolds(1:nrow(dat_for_fa), 2)
dat_efa <- dat_for_fa[idx$Fold1, ]
dat_cfa <- dat_for_fa[idx$Fold1, ]
```

2. Create a scree plot and conduct parallel analysis using your exploratory data set. Based on these results, decide on the number of factors to retain.

```{r echo = FALSE, eval = FALSE}
scree(dat_efa)
paran(dat_efa, cfa = FALSE)
paran(dat_efa, cfa = TRUE)
# Answers may vary - I'll proceed with 4
```

3. Conduct an exploratory factor analysis (using the exploratory data set) with the number of factors that you identified in the previous step. Does your EFA support the theoretical dimensions?

```{r echo = FALSE, eval = FALSE}
efa_res <- fa(dat_efa, nfactors = 4, rotate = "oblimin", fm = "pa")
print(efa_res$loadings, cutoff = .3)
# answers may vary
# not the cleanest solution - 
#  avoidance relates to PA1 and PA3, anxiety mostly relates to PA2
```

4. What changes (if any) would you make to the EFA? Test out any changes you identify.

```{r echo = FALSE, eval = FALSE}
# what if we only use 2 factors?
efa_res <- fa(dat_efa, nfactors = 2, rotate = "oblimin", fm = "pa")
print(efa_res$loadings, cutoff = .3) # somewhat cleaner solution
```

5. Use CFA with the confirmatory data set to test the theoretical structure of these data (outlined at the top of this assignment), and evaluate fit. Does the model fit the data well?

```{r echo = FALSE, eval = FALSE}
mod <- "Avoid =~ RSQ11 + RSQ18 + RSQ21 + RSQ23 + RSQ25
        Anxiety =~ RSQ10 + RSQ12 + RSQ13 + RSQ15 + RSQ20 + RSQ24 + RSQ29 + RSQ30"
cfa_res <- cfa(model = mod, data = dat_cfa, estimator = "MLM")
standardizedSolution(cfa_res)
fitMeasures(cfa_res, 
            fit.measures = c("chisq.scaled", "df.scaled", 
                             "pvalue", "rmsea.scaled", 
                             "srmr", "cfi.scaled", "tli.scaled"))
# yet again, not stellar fit
```

6. What changes (if any) would you consider making to the CFA?

```{r echo = FALSE, eval = FALSE}
# Many possible answers, based on the previous results
# Maybe omit some items or allow cross-loadings based on the EFA results
# Or revisit item text to see if some items don't make theoretical sense
```

