---
title: "Lab 6: Correlation and Regression"
author: "PSYC 6802"
date: "October 8, 2025"
date-format: long
format: 
  revealjs:
    theme: sky
    scrollable: true
    smaller: true
    embed-resources: false
    slide-number: true
    chalkboard: true
    highlight-style: arrow
    link-external-icon: true
execute: 
  echo: true
  freeze: auto
---


```{r include = FALSE}
library(knitr)
opts_chunk$set(class.output='sh', comment = "")
```

## Packges for Today (Install if Needed)

```{r, message = FALSE}
library(tidyverse)
library(pwr)
library(readxl)
library(psych)
```

## Overview

Functions: 

* `cor()` and `cov()`: compute covariance/correlation with arguments:
  + `use`: how to treat missing data, specify "everything", "complete.obs", "pairwise.complete.obs"
  + `method`: which type of correlation, specify "pearson", "kendall", or "spearman"

* `geom_smooth()`: adding best-fit lines to a scatterplot with `ggplot`, with arguments:
  + `method`: `lm` for a linear model or `loess`
  + `formula`: specify the form of a linear model (e.g., quadratic, cubic)
  
* `lm()`: fit a linear regression model

* `scale()`: standardize a variable

* `cor.test()`: hypothesis test for the correlation coefficient ("nil" hypothesis)

* `pwr::pwr.r.test`: power analysis for correlations and simple linear regression

* `poly()`: create polynomial formulas

* `psych::tetrachoric()`: compute tetrachoric correlations

* `psych::biserial()`: compute biserial correlations

## Housing data example

Today, we'll consider the `woodard.xls` dataset:

```{r}
dat <- read_xls("woodard.xls")
```

This data contains information about house features/prices of 100 houses in Raleigh, North Carolina. 

```{r}
dat %>% glimpse
```


## gg-Plotting with Best-Fit Lines

First, we'll look at the relationship between `Building $` and `Sq. Ft.` conditions. 

```{r, fig.height = 4.5}
ggplot(dat, aes(x = `Sq. Ft.`, y = `Building $`)) + 
  geom_point()
```

## gg-Plotting with Best-Fit Lines

We can add different types of best-fit lines using `geom_smooth()`. 

The `method` argument tells `ggplot` which type of best-fit line to draw. The `lm` option indicates a linear model, and the `formula = y ~ x` option indicates simple linear regression.

```{r, fig.height = 3.5}
ggplot(dat, aes(x = `Sq. Ft.`, y = `Building $`)) + 
  geom_point() + 
  geom_smooth(method = lm, formula = y ~ x) 
```

## gg-Plotting with Best-Fit Lines

By default, `ggplot` adds an error band around the best-fit line. We can turn this off using the argument `se = FALSE`.

\footnotesize

```{r, fig.height = 4.5}
ggplot(dat, aes(x = `Sq. Ft.`, y = `Building $`)) + 
  geom_point() + 
  geom_smooth(method = lm, formula = y ~ x, se = FALSE) 
```

## gg-Plotting with Best-Fit Lines

To add the nonparametric loess line, use `method = loess`:

```{r, fig.height = 4.5}
ggplot(dat, aes(x = `Sq. Ft.`, y = `Building $`)) + 
  geom_point() + 
  geom_smooth(method = loess, formula = y ~ x) 
```

## gg-Plotting with Best-Fit Lines

For polynomial best-fit lines, we need to change the `formula` argument. For quadratic regression (a polynomial of degree 2), specify `method = lm` and `formula = y ~ poly(x, 2)`. Here, the method is `lm` because polynomial regression is part of the general linear model.

```{r}
ggplot(dat, aes(x = `Sq. Ft.`, y = `Building $`)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE, 
              formula = y ~ poly(x, 2)) + 
  geom_smooth(method = lm, se = FALSE, 
              formula = y ~ poly(x, 3), col = "red")
```


## Correlations

The `cov()` and `cor()` functions are the easiest ways to calculate the covariance and the correlation.

```{r}
cov(dat$`Year Built`, dat$`Building $`)
cor(dat$`Year Built`, dat$`Building $`)
```

## When `cor()` function does not work :(

Most R functions **will not** handle missing values automatically (it is a good thing). You usually need to tell the function what to do with missing values. For example with `cor()` and `cov()`:

```{r}
cor(dat$`No. Baths`, dat$`Building $`)
```
- the `No. Baths` variable has a missing value, so we need to include the `use =` argument to tell the `cor()` function what to do when missing data is encountered:

```{r}
cor(dat$`No. Baths`, dat$`Building $`, 
    use = "complete.obs")
```
- the `cov()` function works in a similar way

## Correlations: Hypothesis Test

To do a hypothesis test of the Pearson product-moment (PPM) correlation, use the `cor.test()` function:

```{r}
cor.test(dat$`Year Built`, dat$`Building $`)
```

This correlation is significantly different from 0, $r = .29$, $t(98) = 3.01$, $p = .003$, 95% CI[.09, .46]. This function also gives us the confidence interval.

## Correlations: Power Analysis

Finally, we can use the `pwr` package to conduct power analyses, in much the same way as we did for other tests. For example, for an estimated correlation of -0.05. What sample size would be needed to find that this is significantly different than zero with 80% power?

```{r}
pwr.r.test(n = NULL, r = -0.05, sig.level = .05,
           power = .80, alternative = "two.sided")
```

## Correlation Matrices

If you give a matrix of numeric values where the columns represent your variables, the `cor()` function will create a correlation matrix. Here I am entering the first 4 columns of the `dat` object:

```{r}
cor(dat[,1:4])
```


## Other correlation coefficients

For the sake of illustration, let's dichotomize

- `No. Baths` variable to above vs. at/below 2
- `Building $` above and at/below the median price

```{r}
dat <- dat %>%
  mutate(`No. Baths_binary` = as.numeric(`No. Baths`> 2),
         `Building $_binary` = as.numeric(`Building $`> median(`Building $`)))
```

```{r}
table(dat[, c("No. Baths_binary", "Building $_binary")])
```


## Phi Correlations

The correlation between `No. Baths_binary` and `Building $_binary` can be analyzed as either a phi or tetrachoric correlation. Phi correlations are PPMs:

```{r}
cor(dat$`No. Baths_binary`, dat$`Building $_binary`, 
    use = "complete.obs")
```


## Tetrachoric Correlations

Tetrachoric correlations are available from the `psych` package and requires a contingency table as input. 

```{r}
tab <- table(dat$`No. Baths_binary`, dat$`Building $_binary`)
tetrachoric(tab)
```

## (Point-) Biserial Correlations

Biserial correlations treat the binary variable as as dichotom*ized* normal variable. Biserial correlations can be computed using the `biserial` command from the `psych` package. Notice that the dichomized variable must be the *second* variable specified. 

```{r}
cor(dat$`Building $`, dat$`No. Baths_binary`, 
    use = "complete.obs")
biserial(dat$`Building $`, dat$`No. Baths_binary`)
```

## Regression

Finally, let's run a linear regression with `Building $` as the DV and `Sq. Ft.` as the IV. This will give us the equation for the linear best-fit line that we drew on the scatterplot earlier. 

Linear regression in `R` is done with the `lm` function. `lm` takes as input a formula of the form: "DV ~ IV", and you can specify the dataset as the second argument. Usually, you will use the `summary()` function to look at your regression results.

```{r echo = T}
reg <- lm(`Building $` ~ `Sq. Ft.`, dat)
summary(reg)
```

Note that the $t$-test for the regression slope and the $F$-test for $R^2$ is the exact same test in two different forms. The $p$-value is *exactly* the same, and $t^2 = 21.116^2 = 445.9 = F$. 

## Regression

The output tells us that $\widehat{\mathrm{Build.price}} = -52468.43 + 107.18\times \mathrm{Sq.Ft}$. 

* What does the intercept of -52468.43 represent? Is the intercept by itself a useful piece of information in this case?

* For a one-unit increase in `Sq. Ft.`, `Building $` is expected to increase by $107.18 on average.

These interpretations assume that a straight line is an "appropriate" representation of the relationship between our variables.

## Standardized Regression Coefficients

We can find standardized regression coefficients by using `scale()` to standardize each of the variables we input into `lm()`. 

```{r}
summary(lm(scale(`Building $`) ~ scale(`Sq. Ft.`), dat))
```

Is the output what you expect? What pieces are redundant with output we found earlier?

## Regression Assumptions

To be careful about our hypothesis test conclusions, we will want to evaluate our assumptions. With regression, most of our assumptions are about the **residuals**. We can find residuals from the output of `lm()` using the `residuals` function. It can be useful to add these directly to the original data set. 

```{r}
dat <- dat %>% mutate(residuals = residuals(reg))
```

## Evaluting Homoscedasticity

```{r, fig.height=5.5}
ggplot(dat, aes(`Sq. Ft.`, residuals)) + geom_point() + 
  geom_smooth(method = loess, formula = y ~ x) +
  geom_hline(yintercept = 0)
```

## Evaluating Normality

```{r, fig.height = 5.5}
ggplot(dat, aes(sample = residuals)) + 
  stat_qq() + stat_qq_line()
```

