---
title: "Lab 6 Activity"
format: pdf
---

```{r echo = FALSE, eval = FALSE}
library(tidyverse)
```


## 1

In this activity, we will work with the `Prestige` data in the `carData` package.

```{r eval=FALSE, include=TRUE}
data("Prestige", package = "carData")
```


Look at the help page to better understand these data. Here, we want to predict the `income` of occupational incumbents from the average `education` of occupational incumbents. Use `ggplot` to create a scatterplot with one or more best-fit curves. What type of relationship do you think best characterizes the relationship between these variables?

```{r echo = FALSE, eval = FALSE}
data("Prestige", package = "carData")

ggplot(Prestige, aes(education, income)) + 
  geom_point() + geom_smooth(method = lm, formula = 'y ~ x', se = FALSE) +
  geom_smooth(method = loess, formula = 'y ~ x', col = "red", se = FALSE) +
  geom_smooth(method = lm, formula = y ~ poly(x, 2), col = "green", se = FALSE) + 
  geom_smooth(method = lm, formula = y ~ poly(x, 3), col = "purple", se = FALSE)

# quadratic might be a better descriptor than linear
```

## 2

Fit a simple linear regression model predicting `income` from `education` and evaluate assumptions. What do you conclude?

```{r echo = FALSE, eval = FALSE}
mod <- lm(income ~ education, Prestige)
summary(mod)

Prestige <- Prestige %>% mutate(residuals = residuals(mod))

ggplot(Prestige, aes(education, residuals)) + 
  geom_smooth() + geom_point() # probably some heteroscedasticity
ggplot(Prestige, aes(sample = residuals)) + 
  stat_qq() + stat_qq_line() # very non-normal

# there is a statistically significant relationship between education and
#  income (R^2 = .33), but also some major violations of assumptions
```

## 3

Looking ahead: the `lm` function takes formula input like we used with `geom_smooth()`. Fit a regression equation for the best-fit quadratic/cubic curve predicting `income` from `education`. You can find the model residuals in the same way as we did for the linear model. Do the quadratic or cubic models do a better job meeting our assumptions about residuals? If not, can you think of what might be a better solution?

```{r echo = FALSE, eval = FALSE}
mod2 <- lm(income ~ poly(education, 2), Prestige) # quadratic
mod3 <- lm(income ~ poly(education, 3), Prestige) # cubic

summary(mod2)
summary(mod3)
# both models have statistically significant R^2
#  and both have higher R^2 than the linear model

Prestige <- Prestige %>%
  mutate(residuals2 = residuals(mod2),
         residuals3 = residuals(mod3))

ggplot(Prestige, aes(education, residuals2)) + 
  geom_point() + geom_smooth(formula = y ~ x) # still heteroscedastic
ggplot(Prestige, aes(sample = residuals2)) + 
  stat_qq() + stat_qq_line() # still highly non-normal

ggplot(Prestige, aes(education, residuals3)) + 
  geom_point() + geom_smooth(formula = y ~ x) # better, but there are still outliers
ggplot(Prestige, aes(sample = residuals3)) + 
  stat_qq() + stat_qq_line() # still not great

# although the quadratic and cubic models do appear to perform better,
# they still don't meet assumptions well
# a better solution *might* be to log-transform the income variable:

ggplot(Prestige, aes(education, log(income))) + 
  geom_point() + geom_smooth(method = lm)
# although there still are a couple outlying points, this does seem to 
# better meet the regression assumptions
plot(lm(log(income) ~ education, Prestige))
```

