---
title: "Lab 11 Activity"
format: pdf
---


We will work with the same dataset that we used in the main lab. You can read in the data using the following code:

```{r message = FALSE}
library(tidyverse)
library(haven)
dat <- read_dta("Lab_11_attend.dta")
```


## 1

The variables `frosh` and `soph` are dummy variables for the student being a freshman or sophomore (0 on both variables indicates 'other'). Fit a model predicting final exam scores from (the centered) prior GPA with academic year as a moderator (no need to modify the categorical moderator variables). Is there a statistically significant moderation effect?

```{r echo = FALSE, eval = FALSE}
dat <- dat %>% mutate(priGPA_C = priGPA - mean(priGPA), 
                      atndrte_C = atndrte - mean(atndrte))
mod <- lm(final ~ priGPA_C * frosh + priGPA_C * soph, dat)
mod <- lm(final ~ priGPA_C * (frosh + soph), dat) # another way to specify the same model
summary(mod)

# the moderation effect is significant to freshman, but not for sophomores
```

## 2

Construct a plot to illustrate the moderation effect (use separate lines for different grade levels). How does this plot help you interpret your results from step 1?

```{r echo = FALSE, eval = FALSE}
dat_plot <- tibble(priGPA_C = seq(min(dat$priGPA_C), max(dat$priGPA_C)))
dat_plot <- dat_plot %>%
  mutate(frosh = 26.3465 - 0.8439 + priGPA_C * (4.1172 - 2.9060),
         soph = 26.3465 - 0.7571 + priGPA_C * (4.1172 - 0.4441),
         other = 26.3465 + 4.1172 * priGPA_C) %>%
  pivot_longer(-priGPA_C, names_to = "grade", values_to = "final")
ggplot(dat_plot, aes(priGPA_C, final, col = grade)) + geom_path()

## There is a positive (predicted) relationship between prior GPA and
##  final for all grade levels, but this relationship has a less steep
##  slope for freshmen
```

```{r echo = FALSE, eval = FALSE}
## alternative:
library(ggeffects)
ggp <- ggpredict(mod, terms = c("priGPA_C", "frosh", "soph"))
plot(ggp)

```

## 3

Using this same data set, conduct a mediation analysis predicting final exam scores from ACT scores with `hwrte` (rate of completing homework) as a mediator (you may drop the 6 missing values of `hwrte` for this step). What do you conclude? 

```{r echo = FALSE, eval = FALSE}
mod_m <- lm(hwrte ~ ACT, dat)
mod_y <- lm(final ~ ACT + hwrte, dat)


library(mediation)
summary(mediate(model.m = mod_m, model.y = mod_y,
                treat = "ACT", mediator = "hwrte", boot = TRUE))

## The mediation effect is statistically
##  significant, although the effect is rather small
##  (especially compared to the mediator termGPA that we used
##  in lab). Using the effect size measures from the main lecture
##  could help clarify whether this effect is practically significant
```

## 4

Repeat #3 using a method that using a better method of missing data handling than listwise deletion.

```{r echo = FALSE, eval = FALSE}
medmod <- "
final ~ b * hwrte + c * ACT
hwrte ~ a * ACT

ind := a*b    # calculate the indirect effect
tot := a*b+c  # calculate the direct effect
"

fitted <- sem(medmod, data = dat, se = "boot",
              missing = "FIML", estimator = "ML") #, estimator = "MLR") # robust standard errors are not needed with bootstrapping

summary(fitted)

# Defined Parameters:
#                    Estimate  Std.Err  z-value  P(>|z|)
#     ind               0.149    0.027    5.447    0.000
#     tot               0.487    0.048   10.104    0.000
```

```{r echo = FALSE, eval = FALSE}
# brms for the curious
library(mice)
library(brms)
imputed_data <- mice(dat, m = 5)

brms_form <- bf(final ~ hwrte + ACT) + bf(hwrte ~ ACT) + set_rescor(FALSE)

brms_fitted <- brm_multiple(brms_form, data = imputed_data)

summary(brms_fitted)

library(bayestestR) # a package that facilitates mediation analysis with brms
mediation(brms_fitted)
```


