---
title: "Debugging Quarto"
format: docx
---


```{r}
#load data
dat <- read.csv("sat data.csv",
                header = TRUE)
```


**1.** Calculate the standard error for the mean SAT score in Condition C.

```{r message=FALSE, warning=FALSE}
install.packages("dplyr")
library(dplyr)

(se_c <- as.numeric(dat %>% 
                    filter(condition == "c") %>% 
                      summarize(SE = sd(score)/sqrt(nrow(.)))))
```


**2.** Calculate the 95% confidence interval for the mean SAT score in Condition C.

```{r}

n_c <- sum(dat$condition == "c")
mean_c <- mean(dat$score[dat$condition == "c"])

CI_c <- se_c*qt(p = .025, 
                df = n_c - 1, 
                lower.tail = FALSE)

upperCI_c <- mean_c + CI_c  
lowerCI_c <- mean_c - CI_c
```

```{r echo=FALSE}
cat(paste0("95%CI[", round(lowerCI_c, 2), ";", round(upperCI_c, 2), "]"))
```

**3.** Create a plot of boxplots for SAT score by condition,

```{r message=FALSE, warning=FALSE}
dat %>% 
  ggplot(aes(x = condition, y = score)) +
    geom_boxplot(width = .25) + 
    ylim(c(200, 800))
```

**4.** Create Q-Q plots by group for each condition. Make it so the plot matrix has two rows of two plots, instead of one row of four plots.

```{r message=FALSE, warning=FALSE}
dat %>%
  ggplot(aes(sample = score)) +
    geom_qq() +
    geom_qq_line() +
    facet_wrap(. ~ condtion)
```


**5. ** Run Levene’s test. Based on these results, as well as visual inspection of the plots you made, comment on whether the assumptions of normality and homogeneity of variance are violated.

```{r message=FALSE, warning=FALSE}
library(car)

leveneTest(score ~ condition, dat)
```


The assumption of homogeneity of variances has not been violated. However, it certainly seems that the distribution of *group B* presents a noticeable right-skewed, as both the violin plot and QQ plot show  signs of non-normality. In addition, Shapiro-Wilk's test also suggests a violation of the normality assumption for *group B*. Hence, one could be slightly concerned about including *group B* in the final analysis. 

**6.** Find the standardized mean difference between Condition C and Control.

```{r}
library(effectsize)

cohens_d(dat$score[dat$condition == "c"], dat$score[dat$condition == "control"])
```

**7.** Perform a one way ANOVA for score based on condition, and call `summary()`. Then, without using `eta_squared()`, find \eta^2. HINT: you may be able to use some of the values from the `summary()` output as a shortcut. Check your work using `eta_squared()`

```{r}
ANOVA <- aov(score ~ condition, dat)

(AOV_summary <- summary(ANOVA))
```

```{r echo=FALSE}
cat("Hand calculation:")
```

```{r}
(eta_sq <- (AOV_summary[[1]]$`Sum Sq`[1])/sum(AOV_summary[[1]]$`Sum Sq`))
```


```{r echo=FALSE}
cat("Function output:")
```

```{r}
eta_squared(ANOVA)
```



**8.** Use `omega_squared()` to get ω2, then interpret the results of your analysis from questions 8 - 10.


```{r}
omega_squared(ANOVA)
```

The *F*-ratio is significantly different from 0, which implies that the effect of tutoring on SAT results is unlikely to be due to chance. However, both $\eta^2$ and $\omga^2$ suggest that this effect is rather small, with `condition` only accounting for around 4% of the variance in SAT scores. 

# Extra Credit (10 points) 

The expected population mean of SAT scores on the math section is 500, with SD = 100.

-   Find the 95% confidence interval using a normal distribution for a sample of size 5 with mean = 500, SD = 100.

```{r}
se_1 <- 100/sqrt(5)
upperCI_1 <- 500 + (1.96*se_1)
lowerCI_1 <- 500 - (1.96*se_1)

cat(paste0("95%CI[", round(lowerCI_1, 2), ";", round(upperCI_1, 2), "]"))
```

-   Find the 95% confidence interval using a normal distribution for a sample of size 500 with mean = 500, SD = 100.

```{r message=FALSE, warning=FALSE}
se_2 <- 100/sqrt(500)

upperCI_2 <- 500 + (1.96*se_2)

cat(paste0("95%CI[", round(lowerCI_2, 2), ";", round(upperCI_2, 2), "]"))

lowerCI_2 <- 500 - (1.96*se_2)
```

-   Find the 95% confidence interval using a t-distribution, instead of a normal distribution for a sample of size 5 with mean = 500, SD = 100.

```{r}
se_3 <- 100/sqrt(5)
upperCI_3 <- 500 + (qt(p = .025, df = 5 - 1,lower.tail = FALSE)*se_3)
lowerCI_3 <- 500 - (qt(p = .025, df = 5 - 1,lower.tail = FALSE)*se_3)

cat(paste0("95%CI[", round(lowerCI_3, 2), ";", round(upperCI_3, 2), "]"))
```

-   Find the 95% confidence interval using a t-distribution, instead of normal distribution, for a sample of size 500 with mean = 500, SD = 100

```{r}
se_4 <- 100/sqrt(500)
upperCI_4 <- 500 + (qt(p = .025, df = 500 - 1,lower.tail = FALSE)*se_4)
lowerCI_4 <- 500 - (qt(p = .025, df = 500 - 1,lower.tail = FALSE)*se_4)

cat(paste0("95%CI[", round(lowerCI_4, 2), ";", round(upperCI_4, 2), "]"))
```


