---
title: "Lab 5: Power Analysis"
author: "PSYC 6802"
date: "October 1, 2025"
date-format: long
format: 
  revealjs:
    theme: sky
    scrollable: true
    smaller: true
    embed-resources: false
    slide-number: true
    chalkboard: true
    highlight-style: arrow
    link-external-icon: true
    html-math-method: mathjax
execute: 
  echo: true
  freeze: auto
---

## Packages for Today (Install If Needed)

```{r, message = FALSE}
library(tidyverse)
library(pwr)
```


## Overview

The primary package we'll use for conducting power analyses is `pwr`.

Functions we will be using today: 

* `pwr.norm.test`
* `pwr.t.test`
* `pwr.t2n.test`
* `pwr.chisq.test`


## The Logic of the `pwr` Package Functions

Most of the functions of the `pwr` structure work in the same way in that the usually have four main arguments:

- The effect size, `d =`, in the case of *t*-tests.
- The sample size, `n =`
- The type I error rate, `sig.level =`
- The power of your experiment, `power =` 

The `pwr` functions always return the value of the one argument that you leave blank out of these four. 

- These arguments may be named differently depending on the type of power analysis, so always check the function help menu. 

 
## Power for $z$-tests

In the case of a *z*-test, to get **power** given effect size, sample size, and desired type I error rate:

```{r}
pwr.norm.test(d = .5, 
              n = 10, 
              sig.level = .05, 
              power = NULL,
              alternative = "two.sided")
```


## Power for $z$-tests

In the case of a *z*-test, to get required **sample size** given effect size, desired type I error rate, and power:

```{r}
pwr.norm.test(d = .5, 
              n = NULL, 
              sig.level = .05, 
              power = .8,
              alternative = "two.sided")
```

## Power for $z$-tests

The `pwr` package also includes a plotting method. This means that we can use the function `plot()` with the output from any power calculation results. (they don't look very attractive)

```{r, fig.height=4}
plot(pwr.norm.test(d = .5, 
              n = 10, 
              sig.level = .05, 
              power = NULL,
              alternative = "two.sided"))
```


## One-sample $t$-test

In lab 4 we ran a one sample *t*-test to test whether average restaurant ratings were significantly different from 3. with $d = .21$ and $N = 200$ we can do a retrospective power calculation: 

```{r}
pwr.t.test(n = 200, 
           d = .21, 
           sig.level = .05, 
           power = NULL, 
           type = "one.sample", 
           alternative = "two.sided")
```

Our power was around .84. What does this mean?

## One-sample $t$-test

What if I want to repeat the previous "experiment", and I want to know the sample size that I would need to have .8, .9, and .95 power?

```{r, eval = TRUE}
pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .80,
           type = "one.sample", alternative = "two.sided")

pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .90,
           type = "one.sample", alternative = "two.sided")

pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .95,
           type = "one.sample", alternative = "two.sided")
```

For a priori power analysis, we want to round up, so $N = 180, 240, 297$ for 80%, 90%, and 95% power.

Power analysis for a paired-sample *t*-test is the same. 

## Independent-samples $t$-test

For the independent samples $t$-test, we can use `pwr.t.test` if we have equal sample sizes, or `pwr.t2n.test` for unequal sample sizes. 

In the case of lab 4, where we were testing whether males ($n_{1} = 112$) and females ($n_{2} = 82$) were significantly different on average food rating. the estimate effect was $d = .17$, but the difference was not significant $p = .224$:

```{r}
pwr.t2n.test(n1 = 112, n2 = 82, d = .17, sig.level = .05,
             power = NULL, alternative = "two.sided")
```

In retrospect, their power to detect an effect was only .21! 

## Independent-samples $t$-test

When planning data collection, you ideally want to plan for groups to have an equal size if possible. So, we can calculate a priori power assuming equal sample sizes and $d = .17$ as an effect size estimate. 

```{r, eval = TRUE}
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .80,
           type = "two.sample", alternative = "two.sided")
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .90,
           type = "two.sample", alternative = "two.sided")
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .95,
           type = "two.sample", alternative = "two.sided")

```

It turns out that $n$ = 545, 729, and 901 (quite large!) are the minimum sample sizes **in each group** for 80%, 90%, and 95% power.

## Power for $\chi^2$ tests

We can do the same for $\chi^2$ tests using the `pwr.chi.test` function. The test does not distinguish between $\chi^2$ goodness-of-fit and $\chi^2$ test of independence. 

Let's use the airplane delays example from Lab 3. $\chi^2(1) = 410.66$ with $N = 106596$. Our $W$ coefficient was $.06$. 

```{r}
pwr.chisq.test(w = .06, N = 106596, df = 1, sig.level = .05, power = NULL)
```

Power is essentially 1, we are guaranteed to always reject the null hypothesis under these circumstances. 

## Power for $\chi^2$ tests

Next, let's conduct a priori power for this study.

```{r}
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .80)
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .90)
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .95)
```

For this effect size, we would just need $N = 2181, 2919, 3610$ for 80%, 90%, and 95% power.


# My Quarto file Won't Compile :(

## Let's work together

On Blackboard, there is a Quarto file that contains errors that will cause it not to render. Download the `Debugging_Quarto.qmd` file and let's try to solve the errors together until the file renders!

- We will also need the `sat_data.csv` dataset to run the code. 

- Don't worry too much about understanding the *content* of the document, focus on trying to get the output to render and look as it should

The following slides are not very detailed, as the process of solving rendering issues varies greatly.


## Step 1: Make sure R code works

Before you even attempt to render the document, it is good practice to clean your environment and run all your R code chunks sequentially.

- To quickly run all the code chunks either navigate to the `run` option in the to right of the source pane and click `Run All`, or use the shortcut `CMD + Option + R` on MAC or `CTRL + Alt + R` on Windows

Let's solve the code errors until all the chunks run sequentially. 

## Step 2: Render and solve LaTeX errors

If all your R chunks run fine and your document still does not render, there is probably some problem with other parts of your text. 

Read the error messages, but know that some messages are more helpful than others.

* Some error messages reference lines in your .qmd file

* Some error messages reference lines in a .tex or .log file

If your Quarto file fails to render, you may find a **.tex** file and/or a **.log** file in the same folder as the .qmd document.

* Don't edit the .tex file directly, but inspecting this file may give you more information about what causes the problem

## Some More Debugging Strategies

* Pay attention to which file that the error message references - line numbers will be the line number in that file

* Do an internet search for the error message

* Run smaller parts of the document to help diagnose where the problem lies (delete or comment out other sections)

* **Persevere!** Take breaks, give yourself enough time to deal with these potential logistical problems. Debugging often takes longer than you think that it should