---
title: "Lab 5: Power Analysis"
---

## Packages for Today (Install If Needed)

```{r, message = FALSE}
library(tidyverse)
library(pwr)
```


## Overview

The primary package we'll use for conducting power analyses is `pwr`.

Functions we will be using today: 

* `pwr.norm.test`
* `pwr.t.test`
* `pwr.t2n.test`
* `pwr.chisq.test`


## The Logic of the `pwr` Package Functions

Most of the functions of the `pwr` structure work in the same way in that the usually have four main arguments:

- The effect size, `d =`, in the case of *t*-tests.
- The sample size, `n =`
- The type I error rate, `sig.level =`
- The power of your experiment, `power =` 

The `pwr` functions always return the value of the one argument that you leave blank out of these four. 

- These arguments may be named differently depending on the type of power analysis, so always check the function help menu. 

 
## Power for $z$-tests

In the case of a *z*-test, to get **power** given effect size, sample size, and desired type I error rate:

```{r}
pwr.norm.test(d = .5, 
              n = 10, 
              sig.level = .05, 
              power = NULL,
              alternative = "two.sided")
```


## Power for $z$-tests

In the case of a *z*-test, to get required **sample size** given effect size, desired type I error rate, and power:

```{r}
pwr.norm.test(d = .5, 
              n = NULL, 
              sig.level = .05, 
              power = .8,
              alternative = "two.sided")
```

## Power for $z$-tests

The `pwr` package also includes a plotting method. This means that we can use the function `plot()` with the output from any power calculation results. (they don't look very attractive)

```{r, fig.height=4}
plot(pwr.norm.test(d = .5, 
              n = 10, 
              sig.level = .05, 
              power = NULL,
              alternative = "two.sided"))
```


## One-sample $t$-test

In lab 4 we ran a one sample *t*-test to test whether average restaurant ratings were significantly different from 3. with $d = .21$ and $N = 200$ we can do a retrospective power calculation: 

```{r}
pwr.t.test(n = 200, 
           d = .21, 
           sig.level = .05, 
           power = NULL, 
           type = "one.sample", 
           alternative = "two.sided")
```

Our power was around .84. What does this mean?

## One-sample $t$-test

What if I want to repeat the previous "experiment", and I want to know the sample size that I would need to have .8, .9, and .95 power?

```{r, eval = TRUE}
pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .80,
           type = "one.sample", alternative = "two.sided")

pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .90,
           type = "one.sample", alternative = "two.sided")

pwr.t.test(n = NULL, d = .21, sig.level = .05, power = .95,
           type = "one.sample", alternative = "two.sided")
```

For a priori power analysis, we want to round up, so $N = 180, 240, 297$ for 80%, 90%, and 95% power.

Power analysis for a paired-sample *t*-test is the same. 

## Independent-samples $t$-test

For the independent samples $t$-test, we can use `pwr.t.test` if we have equal sample sizes, or `pwr.t2n.test` for unequal sample sizes. 

In the case of lab 4, where we were testing whether males ($n_{1} = 112$) and females ($n_{2} = 82$) were significantly different on average food rating. the estimate effect was $d = .17$, but the difference was not significant $p = .224$:

```{r}
pwr.t2n.test(n1 = 112, n2 = 82, d = .17, sig.level = .05,
             power = NULL, alternative = "two.sided")
```

In retrospect, their power to detect an effect was only .21! 

## Independent-samples $t$-test

When planning data collection, you ideally want to plan for groups to have an equal size if possible. So, we can calculate a priori power assuming equal sample sizes and $d = .17$ as an effect size estimate. 

```{r, eval = TRUE}
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .80,
           type = "two.sample", alternative = "two.sided")
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .90,
           type = "two.sample", alternative = "two.sided")
pwr.t.test(n = NULL, d = .17, sig.level = .05, power = .95,
           type = "two.sample", alternative = "two.sided")

```

It turns out that $n$ = 545, 729, and 901 (quite large!) are the minimum sample sizes **in each group** for 80%, 90%, and 95% power.

## Power for $\chi^2$ tests

We can do the same for $\chi^2$ tests using the `pwr.chi.test` function. The test does not distinguish between $\chi^2$ goodness-of-fit and $\chi^2$ test of independence. 

Let's use the airplane delays example from Lab 3. $\chi^2(1) = 410.66$ with $N = 106596$. Our $W$ coefficient was $.06$. 

```{r}
pwr.chisq.test(w = .06, N = 106596, df = 1, sig.level = .05, power = NULL)
```

Power is essentially 1, we are guaranteed to always reject the null hypothesis under these circumstances. 

## Power for $\chi^2$ tests

Next, let's conduct a priori power for this study.

```{r}
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .80)
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .90)
pwr.chisq.test(w = .06, N = NULL, df = 1, 
               sig.level = .05, power = .95)
```

For this effect size, we would just need $N = 2181, 2919, 3610$ for 80%, 90%, and 95% power.

