---
title: "Lab 7 Activity"
format: pdf
---

## 1

Load in the `PlantGrowth` data from the `datasets` package. You can do so by running:

```{r}
data("PlantGrowth", package = "datasets")
```

These are the data discussed in the first part of the one-way ANOVA lecture. The IV is `group` and the DV is `weight`.

## 2

Test the assumptions of your ANOVA using the methods we learned in earlier weeks to test assumptions about normality and outliers.

```{r echo = TRUE, eval = TRUE}
library(tidyverse)
## outliers
# from slide 4 in the lecture slides, trt1 might have some outliers
PlantGrowth %>% group_by(group) %>%
  summarize(mean = mean(weight), tmean = mean(weight, trim = .1), .groups = "keep")
# no highly influential outliers here

## normality
PlantGrowth %>% group_by(group) %>%
  summarize(skew = e1071::skewness(weight), 
            kurt = e1071::kurtosis(weight), .groups = "keep") 
# potentially significant skew and kurtosis
ggplot(PlantGrowth, aes(sample = weight)) + 
  stat_qq() + stat_qq_line() + facet_wrap(~group)

# possibly some deviations from normality, but
#  n is so small that it's hard to tell
# Robust tests are more advisable
```


## 3 

Run an ANOVA that is appropriate for these data using the package of your choice, and compute an effect size. 

```{r echo = TRUE, eval = TRUE}
# a number of options available, a few are shown below

data("PlantGrowth", package = "datasets")

# aov without weighted least squares
summary(aov(weight ~ group, PlantGrowth))

# aov with weighted least squares

PlantGrowth <- PlantGrowth %>% group_by(group) %>%
  mutate(weights = 1 / var(weight)) %>% ungroup
summary(aov(weight ~ group, PlantGrowth, weights = weights))

# aov with White's adjustment
car::Anova(aov(weight ~ group, PlantGrowth), white.adjust = TRUE)

# ezANOVA - need to add an ID variable
library(ez)
PlantGrowth <- PlantGrowth %>% mutate(ID = 1:nrow(PlantGrowth))
# ezANOVA 
ezANOVA(data = PlantGrowth, dv = weight, wid = ID, between = group) # note that Levene's test NOT rejected

# oneway with Welch's correction:
library(rstatix)
welch_anova_test(PlantGrowth, weight ~ group)


library(effectsize)

res <- aov(weight ~ group, PlantGrowth)

eta_squared(res)
omega_squared(res)

```

## 4

Identify an *a priori* contrast to test on the `PlantGrowth` data, and test it in `R`.

```{r echo = TRUE, eval = TRUE}
library(emmeans)

# one option is to compare the control group to the average of the two treatments
summary(PlantGrowth$group) # factor levels are ordered ctrl, trt1, trt2

aov_res <- aov(weight ~ group, PlantGrowth)
em1 <- emmeans(aov_res, specs = "group")

contrast(em1, method = list(ctr_v_trt = c(2, -1, -1)))
# not statistically significant, but bear in mind that n is small
```

## 5

Conduct post-hoc tests of your choosing on the `PlantGrowth` data. What groups are significantly different from each other with your choice of adjustment for multiple comparisons? In real life, would you recommend conducting these post-hoc tests, and why?

```{r echo = TRUE, eval = TRUE}
# yes, it's appropriate to look at these post-hoc tests since the original ANOVA
# is significant at alpha = .05

# any package is fine
# most methods show that the major difference is between trt1 and trt2

em_res <- emmeans(aov_res, specs = "group")
pairs(em_res, adjust = "holm")
pairs(em_res, adjust = "bonferroni")
pairs(em_res, adjust = "tukey")
```


