---
title: "Lab 7: One-Way ANOVA"
author: "PSYC 6802"
date: "October 22, 2025"
date-format: long
format: 
  revealjs:
    theme: sky
    scrollable: true
    smaller: true
    embed-resources: false
    slide-number: true
    chalkboard: true
    highlight-style: arrow
    link-external-icon: true
execute: 
  echo: true
  freeze: auto
---


```{r include = FALSE}
library(knitr)
opts_chunk$set(class.output='sh', comment = "")
```

## Packages for Today (Install if Needed)

```{r, message = FALSE}
library(tidyverse)
library(ez)
library(pwr)
library(rstatix)
library(emmeans)
library(effectsize)
```

## Today's Data Set

In lecture, we looked at whether *everyday people* rated Drama, Comedy, and Sci-Fi movies differently on Rotten Tomatoes. 

Today, we'll look at the same data, but focus on *professional critics'* ratings, which are availble in the `tomatoMeter` column of the dataset.


## Read in and Clean Data

```{r message=FALSE, warning=FALSE}
dat <- read_csv("rotten_tomatoes_clean.csv")
dat %>% glimpse
```

We will recode our IV `genre` as a factor variable rather than as a character variable. More broadly, this recoding is **very important**, especially when your IV is coded as a numeric variable. If you specify the IV as numeric, most ANOVA functions will give you output, but it will be incorrect. Specifically, these functions will treat group codes as a continuous variable rather than as a grouping variable.

```{r}
dat <- dat %>% mutate(genre = factor(genre))
```

## Visualize Data

```{r}
dat %>% 
  ggplot(aes(genre, tomatoMeter)) + geom_boxplot()
```
No obvious outliers or heteroskedasticity (boxes are about the same size)

# Visualize Data

```{r}
dat %>%
  ggplot(aes(sample = tomatoMeter)) + geom_qq() + geom_qq_line() + 
           facet_wrap(~genre)
```

Each genre seems to have some amount of non-normality


## `aov` Approach

`aov` is `R`'s built-in command for running ANOVAs. `aov` is also the class of object outputted by a function call to `aov()`. This approach assumes homoskedasticity, which we'll address later.

```{r}
aov_res <- aov(tomatoMeter ~ genre, dat)
summary(aov_res)
```

This is very similar to the `lm` approach:

```{r}
anova(lm(tomatoMeter ~ genre, dat))
```


Here, we reject the null hypothesis, indicating that critics rate different movie genres differently on average.


## Effect sizes of Variance Explained: $\eta^2$ and $\omega^2$

The `effectsize` package includes functions to calculate $\eta^2$and $\omega^2$.

```{r}
library(effectsize)
```

::: columns
:::: column

- $\eta^2$

```{r message=FALSE, warning=FALSE}
eta_squared(aov_res)
```


::::
:::: column

- and $\omega^2$, which is always going to be equal to or smaller than $\eta^2$

```{r message=FALSE, warning=FALSE}
omega_squared(aov_res)
```

  
::::
:::


Both of these effects are relatively small


## APA Style Report

A one-way ANOVA was conducted to assess the relationship between the Rotten Tomatoes tomato meter (DV) and movie genre (IV). There was a significant association, $F(2, 1907) = 55.1, p = <.001$, $\omega^2 = .05$.

## The `emmeans` Package

In lecture we showed some "built-in" functions for multiple comparisons. Here we'll look into another option: the `emmeans` package. 

- First you need to create a new object using the `emmeans()` function. The main arguments of this function are you `aov` object and your IV(s), which you specify after the `specs =` argument. 

```{r}
library(emmeans)
em <- emmeans(aov_res, specs = "genre")
```

Then, you can apply many of the `emmeans` package functions to the newly created object to compute all sorts of pairwise comparisons and contrasts.

## Cohen's $d$  with `emmeans`

We can calculate the Cohen's $d$ between every group pairs. If you remember the formula is $d_{i,j} = \frac{M_{i} - M_j}{\sqrt{MSW}}$. The `eff_size()` function does this for all our groups.

- `eff_size()` requires the `em` object, the standard error $\sqrt{MSW}$, and the residual degrees of freedom $df_{\mathrm{within}}$

```{r}
eff_size(em, 
         
         # sigma(aov_res) equals sqrt(MSW)
         sigma = sigma(aov_res), 
         
         # df.residual(aov_res) extracts the df within from the model
         edf = df.residual(aov_res))
```


## Pairwise Comparisons (post-hoc): Tukey's HSD

The `pairs()` function will run pairwise comparisons if you enter the `em` object. You also need to specify the type of adjustment through the `adjust =` argument.

- For example, Tukey's HSD:

```{r}
pairs(em, adjust = "tukey")
```

Here we see that dramas are rated significantly higher than both comedies and sci-fi movies, and that sci-fi is rated significantly higher than comedy

## Post-Hoc Tests with `aov`

There are many types of adjustments, and the `adjust =` argument allows for several:


::: columns
:::: column

- `"tukey"`: this will perform Tukey's HSD.
- `"holm"`: pairwise $t$-tests with Holm-adjusted $p$-values.
- `"bonferroni`: pairwise $t$-tests with Bonferroni-adjusted $p$-values.


::::
:::: column
- `"Dunnett"`: an option that doesn't assume homogeneity. 
- `"none"`: no correction. 

::::
:::

You can try substituting these into the `adjust =` argument from the previous slide; the only thing that will change is the *p*-value. 


## Contrasts

To check contrasts, we can use the `contrast()` function on the object. We can then specify contrasts using `method = `. Note that we must specify the contrasts as a `list`. 

We also need to make sure we reference the correct order of our factor levels. We can check this as follows

```{r}
levels(dat$genre)
```

So what will my contrasts look like if I want to test the hypothesis:

* Drama vs. Comedy/SciFi (combined)

## Contrasts


```{r}
contrast(em, method = list(genre = c(-1, 2, -1)))
```

Here, we reject $H_0$ for both cases, so we find that dramas are rated significantly differently than sci-fi and comedy movies. 


## Weighted Least Squares with `aov`

First, we need to compute weights for WLS estimation. Weights are the inverse of the group-level variances ($W_{\mathrm{group}} = \frac{1}{S^2_{\mathrm{group}}}$). 

```{r}
dat <- dat %>% group_by(genre) %>%
  mutate(weights = 1 / var(tomatoMeter))
```

Next, we add these weights to our `aov()` call.

```{r}
summary(aov(tomatoMeter ~ genre, dat, weights = weights))
```

## White's Adjustment with `aov`

We can do White's adjustment with `aov` output, with a little help from the `car` package.

```{r}
car::Anova(aov_res, white.adjust = TRUE)
```

## White's Adjustment with `aov`

Notice that WLS and White's adjustment give very similar $p$-values.

The message "Coefficient covariances computed by hccm()" refers to a particular version of White's adjustment. We won't go into details of this in this class. 


# Alternative ANOVA Functions and Robust Methods

## Other Packages for ANOVA

The default `R` methods don't include all of the robust estimation methods and post-hoc comparisons that we might want to use. Other packages have been developed to make these methods more easily available and more user-friendly. We'll learn about two ANOVA commands in new packages today: `ez` from `ezANOVA` and `welch_anova_test` from `rstatix`. 

`ezANOVA` makes it easy to specify more complicated models, and it will be the main function we use for more complex ANOVA models. Another advantage of `ezANOVA` is that there is an option to return an `aov` class object, so all of the functions above that used `aov()` objects can also use `ezANOVA` output.

## ANOVA with `ezANOVA`

```{r, eval = FALSE}
# If there is no id column in the original dataset
# you can do 
# mutate(ID = 1:nrow(dat))

ezANOVA(data = dat, dv = tomatoMeter, wid = id, between = genre)
```

Input:

* `data`: dataset (other inputs are column names from this dataset)
* `dv`: dependent variable
* `wid`: within-subjects ID (person identifier variable - required)
* `between`: between-subjects factor (grouping variable in one-way ANOVA)

Output:

* Rather than the full ANOVA table, `ezANOVA` only gives the numerator and denominator $df$, $F$ statistic, $p$ value, and ges = eta-squared effect size ($\eta^2$)
* `ezANOVA` also returns Levene's test for homogeneity of variance. If this test is rejected, you have evidence against homogeneity, and you definitely want to use Welch's test, WLS or White's adjustment.

## ANOVA with `ezANOVA`

```{r}
ezANOVA(data = dat, dv = tomatoMeter, wid = id, between = genre)
```

`ezANOVA` tends to give a lot of warnings, many of which can be ignored. 

* `ezANOVA` will automatically convert your subject identifier variable to a factor (if it not already a factor). This is usually appropriate, so the warning can be ignored. 

* `ezANOVA` will also warn if you don't have equal group sizes. For the one-way ANOVA, you don't need to worry about this (changing the `type` argument won't change your output). This will matter when we have more than one IV.

## ANOVA with `ezANOVA`

To return an `aov` class of object, run:

```{r, warning = FALSE}
ez_res <- ezANOVA(data = dat, dv = tomatoMeter, wid = id, 
                  between = genre, return_aov = TRUE)

ez_res$aov # extract aov output
```

Be careful with this option - not all specifications (e.g., White's adjustment) will be carried over to the `aov` object. 

## Welch's test

Welch's correction is available in the `rstatix` package. Functions in this package will not run with grouped data (we used the `group_by` function earlier), so we ungroup the data first.

```{r}
dat <- dat %>% ungroup

welch_anova_test(dat, tomatoMeter ~ genre)
```

## Kruskal-Wallis Rank Sum Test

Finally, if we wish to avoid assumptions of normality, we can run the Kruskal-Wallis rank sum test, which also rejects the null hypothesis.

```{r}
kruskal.test(tomatoMeter ~ genre, dat)
```
To follow up with multiple comparisons, we might use pairwise tests that make the same assumptions, such as the pairwise Wilcoxon rank sum test:

```{r}
pairwise.wilcox.test(x = dat$tomatoMeter, g = dat$genre, p.adjust.method = "holm")
```



# Summary 

## Summary of Useful Functions:

Many new functions today! Here is a summary of all of the functions we used and what package they belong to.

Functions:

* Fit a one-way ANOVA
  - `aov(DV ~ IV, data)`  (returns `aov` class output)
  - `anova(lm(DV ~ IV, data))` (very similar to `aov()`)
  - `ez::ezANOVA(data, DV, ID, between = IV)` (in `ez`) 
  - `ez::ezANOVA(data, DV, ID, between = IV, return_aov = TRUE)` (returns `aov` output)

* Heterogeneity of Variance Corrections
  - `rstatix::welch_anova_test(dataset, IV ~ DV)` (Welch)
  - `ez::ezANOVA(data, DV, ID, between = IV, white.adjust = TRUE)` (White)
  - `car::Anova(aov(DV ~ IV, data), white.adjust = TRUE)` (White)
  - `aov(DV ~ IV, data, weights = ...)` (WLS)
  - `anova(lm(DV ~ IV, data, weights = ...))` (WLS)

## Summary of Useful Functions (cont.):

* Effect Sizes
  - `effectsize::eta_squared(aov)` ($\eta^2$)
  - `effectsize::omega_squared(aov)` ($\omega^2$)
  - `emmeans::eff_size(emmGrid, sigma, edf)`(Cohen's *d*)
  - `ez::ezANOVA(data, DV, ID, between = IV)` ($\eta^2$)
  
* Power Analysis (covered in lecture slides, not lab)
  - `pwr::pwr.anova.test()` (in `pwr`, equal sized groups)
  - `pwr::pwr.f2.test()` (in `pwr`, groups need not be same size)
  

## Summary of Useful Functions (cont.):

* Post-Hoc Tests
  - `emmeans::pairs(emmeans(aov, specs = IV, adjust = ...))`

* A Priori Contrasts (covered in lecture slides, not lab)
  - `emmeans::contrast(emmeans(aov, specs = IV), method = list(= c(...)))`
  - `summary(multcomp::glht(aov, mcp(DV = c(...))))`

* Nonparametric Methods

  - `kruskal.test(DV ~ IV, data)`
  - `pairwise.wilcox.test(x = DV, g = IV, p.adjust.method = ...)`