---
title: "Lab 8: Factorial ANOVA"
author: "PSYC 6802"
date: October 29, 2025
date-format: long
format: 
  revealjs:
    theme: sky
    scrollable: true
    smaller: true
    embed-resources: true
    slide-number: true
execute: 
  echo: true
---

## Packages for Today

```{r, message = FALSE}
library(tidyverse)
library(rio)
library(effectsize)
library(labelled)
library(car)
library(ez)
library(sjstats)
library(emmeans)
library(ggeffects)
theme_set(theme_classic())
```

## Data for Today

Rotten Tomatoes Data

* IVs: Genre (Comedy vs. Drama vs. Sci-Fi) and Rating (PG vs. PG-13 vs. R)

* DV: Box office revenue in millions of US dollars

```{r}
dat <- read_csv("rotten_tomatoes_clean.csv")

dat %>% summary
```

## Data Cleaning

The DV of box office revenue (`boxOfficeMil`) is missing for two movies, so we will remove these from the data set:

```{r}
dat <- dat %>% filter(!is.na(boxOfficeMil))
```

We also want to omit movies without the PG, PG-13, or R ratings:

```{r}
dat <- dat %>% filter(rating %in% c("PG", "PG-13", "R"))
```

Check for a balanced design (equal sample sizes for each combination of groups):

```{r}
dat %>% select(rating, genre) %>% table
```
## Box Plots

Let's first explore the distribution of these data using box plots:

```{r}
dat %>%
  ggplot(aes(rating, boxOfficeMil, color = genre)) + 
  geom_boxplot()
```

## Data Transformation

We see (as we might have anticipated), a large number of positive outliers. This is an instance in which we might prefer to analyze a transformed outcome variable. This time, we'll use the Box-Cox transformation from Lecture 1.

```{r}
car::powerTransform(dat$boxOfficeMil)

dat <- dat %>% mutate(boxOffice_transformed = bcPower(boxOfficeMil, lambda = 0.2164347))
```

* With this transformation, we lose the ability to easily interpret the metric of the DV in favor of better meeting our test assumptions.

* Be sure to always apply the exact same transformation to all values (no group-level transformations!)


## Box Plots

```{r}
dat %>%
  ggplot(aes(rating, boxOffice_transformed, color = genre)) + 
  geom_boxplot()
```

Now, we still have a few outliers, but our data should better meet the normality assumption with the transformed data.


## Interaction Plot

The easiest way to make an interaction plot with `ggplot` is to create a new data set that includes group means.

```{r}
dat_means <- dat %>% 
  group_by(rating, genre) %>%
  summarize(mean_boxOffice_transformed = mean(boxOffice_transformed), .groups = "keep")
```

## Interaction Plot

Then, we can use this to draw an interaction plot.

```{r}
ggplot(dat_means, 
       aes(rating, mean_boxOffice_transformed,
           group = genre, 
           col = genre)) + 
  geom_point() + geom_line()
```

## Formulas in R

Earlier, we saw the `DV ~ IV` notation for formulas.

We can also use `DV ~ IV1 * IV2` to include both main effects and an interaction. 

* A longer way to write the same formula is `DV ~ IV1 + IV2 + IV1:IV2`.

* The formula `DV ~ IV1:IV2` includes only the interaction, not the main effects.

* The "*" notation implies all lower-order terms, and the ":" notation implies only the interaction effect. 

## Two-Way ANOVA with `aov`

**Warning**: Don't try this at home if you have unbalanced data!

* If data are unbalanced, the main effects results depend on the order of the variables. Compare the two solutions:


::: {.panel-tabset}

## Rating * Genre

```{r}
aov_res <- aov(boxOffice_transformed ~ rating * genre, dat)
summary(aov_res)
```

## Genre * Rating

```{r}
aov_res <- aov(boxOffice_transformed ~ genre * rating, dat)
summary(aov_res)
```
:::


## Two-Way ANOVA with `ezANOVA`

`ezANOVA` in the `ez` package allows us to choose the sums of squares type. Here, we'll use type 3 SS since we are interested in a potential interaction effect. 


```{r}
ezANOVA(data = dat, dv = boxOffice_transformed,
        between = c(genre, rating), wid = id,
        type = 3)
```


## Two-Way ANOVA with `ezANOVA`

We rejected the null hypothesis for Levene's test, so we should use a correction for violations of homogeneity of variance. One option is to use White's adjustment, which is available with `ezANOVA`.

```{r}
ezANOVA(data = dat, dv = boxOffice_transformed,
        between = c(genre, rating), wid = id,
        type = 3, white.adjust = TRUE)
```

## Two-Way ANOVA with `aov`

Another option if homogeneity is violated is to use weighted least squares. We can calculate and specify weights with the `aov` function.

Calculate weights for each combination of groups:

```{r}
dat <- dat %>% 
  group_by(genre, rating) %>%
  mutate(weight = 1 / var(boxOffice_transformed))
```

Run WLS ANOVA:

```{r}
res_aov <- aov(boxOffice_transformed ~ genre * rating, 
               dat, weights = weight)

```

Because we have unbalanced data, we want to print type 2 or 3 sums of squares. Below, we find type 2 sums of squares using the `Anova` function from the `car` package. Note that we do **not** want to use this function to find type 3 sums of squares (use `ezAnova` instead).

```{r}
Anova(res_aov, type = 2)
```

## $r$-Family Effect Sizes 

Now that we have more effects in our ANOVA model, calculating effect size becomes more nuanced. **Partial** effect sizes enter the picture.

- normally, $\eta^2$ and $\omega^2$ compare the variance explained of one of the ANOVA effects against the total variance.
- *partial* $\eta^2$ and $\omega^2$ compare the variance explained of one of the ANOVA effects against the total variance *after* removing the variance already explained by other terms. 

In practice, the numerator of the formula remains the same, but the denominator is always smaller for partial measures of effect sizes. This means that partial measures are *always* larger than their counterparts. 

## Effect Sizes 

The `sjstats` package includes the `anova_stats` function for calculating $r$-family effect sizes. This is especially useful when you have multiple main effects and interaction effects:

```{r}
anova_stats(res_aov)
```

For the interaction effect, $\eta^2 = .006$, partial $\eta^2 = .007$, $\omega^2 = .004$, and partial $\omega^2 = .005$.

## Simple Main Effects by Rating Category

We found that the interaction effect is statistically significant at $\alpha = .05$. To better understand what this result means, let's calculate simple main effects. You should *always* do this whenever an interaction in your model is significant. (If the interaction is **not** significant, look at the ordinary main effects in the ANOVA table instead.)

Let's split the data by the three ratings, then look at the effects of genre on box office revenue within each rating category.

```{r}
dat_pg <- dat %>% filter(rating == "PG")
dat_pg13 <- dat %>% filter(rating == "PG-13")
dat_r <- dat %>% filter(rating == "R")
```

Next, we want the $MSW$ and $dfW$ from the original model. From before, $MSW = 9.8$ and $dfW = 1895$. (Note that the type of SS doesn't affect these numbers.)

## Simple Main Effects by Rating Category

Next, conduct a one-way ANOVA for each sub-data set.

```{r}
summary(aov(boxOffice_transformed ~ genre, dat_pg))
summary(aov(boxOffice_transformed ~ genre, dat_pg13))
summary(aov(boxOffice_transformed ~ genre, dat_r))
```

* For the PG rated movies, $MSB = 147.64$ and $dfB = 2$.

* For the PG-13 rated movies, $MSB = 552.1$ and $dfB = 2$.

* For the R rated movies, $MSB = 220.13$ and $dfB = 2$.

## Simple Main Effects by Rating Category

The $F$ test statistic equals $MSB/MSW$, where $MSB$ is from the sub-data sets, and $MSW$ is from the original ANOVA.

```{r}
F_pg <- 147.64 / 9.8
F_pg13 <- 552.1 / 9.8
F_r <- 220.13 / 9.8
```

To find $p$-values, find the proportion of the $F$ distribution above each $F$ test statistic. Be sure to use $dfB$ from the sub-dataset and $dfW$ from the original dataset.

```{r}
pf(F_pg, df1 = 2, df2 = 1895, lower.tail = FALSE)
pf(F_pg13, df1 = 2, df2 = 1895, lower.tail = FALSE)
pf(F_r, df1 = 2, df2 = 1895, lower.tail = FALSE)
```

At $\alpha = .05$, there are significant differences in genre at all three rating categories.


## Simple Main Effects Using `joint_tests()`

An easier way to do these tests imply is to use the `joint_tests()` function from the `emmeans` package. First we define an object using the `emmeans()` function with both grouping variables under `specs`


```{r}
em1 <- emmeans(aov_res, specs = c("genre","rating"))
```

Then we can use `joint_tests()` to test whether the effect of `rating` is significant `by` the different levels of `rating``

```{r}
joint_tests(em1, by = "rating")
```

The problem with this approach is we get much less detail in the output, some of which we need to compute effect sizes.

## Post-Hoc Tests for Simple Main Effects

If there are 3 or more levels of the IV for simple effects, we can do post-hoc or a priori analyses like we learned about for the one-way ANOVA. For example, we can do Tukey's HSD for each rating separately.

```{r}
pairs(em1, by = "rating", adjust = "tukey") 
```
All pairwise comparisons are statistically significant except for comedy and sci-fi movies with a PG rating.

## Effect Sizes for Simple Main Effects

We can use the equations from lecture and `R` as a calculator to find effect sizes for the simple main effect of genre separately for each rating category. 

$$Partial\ \eta^2_{simple} = \frac{SSB_{simple}}{SSB_{simple} + SSW}$$
$$\eta^2_{simple} = \frac{SSB_{simple}}{SST}$$

$$Partial\ \omega^2_{simple} = \frac{dfB(MSB_{simple} - MSW)}{SSB_{simple} + (N - dfB)MSW}$$
$$\omega^2_{simple} = \frac{dfB(MSB_{simple} - MSW)}{SST + MSW}$$

## Calculations: PG Example

```{r}
summary(aov(boxOffice_transformed ~ genre, dat_pg))
```

$N 326$, $SS_{simple} = 295.3$, $MS_{simple} = 147.64$ and $df_{simple} = 2$ from the sub-data set. 

```{r}
summary(aov(boxOffice_transformed ~ genre * rating, dat))
```


$SS_{W} = 18484$, $df_{W} = 1895$, $MS_{W} = 9.8$, and $SS_{T} = 2938 + 1759 + 140 + 18484 = 23321$

```{r}
295.3 / 23321 # eta-squared
295.3 / (295.3 + 18484) # partial eta-squared
2 * (147.64 - 9.8) / (23321 + 9.8) # omega-squared
2 * (147.64 - 9.8) / (295.3 + (326 - 2) * 9.8)  # partial omega-squared
```
