---
title: "Lab 8 Activity"
format: pdf
---

## 1

Load in the `Ornstein` data from the `carData` package, along with the `tidyverse` package. Read the help page for the `Ornstein` data to better understand these data.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
data("Ornstein", package = "carData")
```

For today, we will use `interlocks` as the DV, and `sector` and `nation` as IVs. However, we are only interested in whether the nation is Canada or not, so we want to combine the `OTH`, `UK`, and `US` factor levels. One way to accomplish this is to run the following code. (Then use `canada` as an IV instead of `nation`).

```{r eval = TRUE}
Ornstein <- Ornstein %>%
  mutate(canada = factor(nation == "CAN"))
```

Finally, we will omit data from the "BNK", "CON", and "HLD" levels of `sector` because they have relatively few observations.

```{r eval = TRUE}
Ornstein <- Ornstein %>%
  filter(!sector %in% c("BNK", "CON", "HLD"), .preserve = FALSE) %>%
  mutate(sector = factor(sector))
```


## 2

Conduct graphical and numerical exploratory analyses on these data. How do these analyses inform how you set up the ANOVA and what results you expect to find?

```{r echo = TRUE, eval = TRUE}
# quantitative: we want to check for balanced data among our IVs
table(Ornstein$sector, Ornstein$canada) # unbalanced data - should use type II or III SS

# graphical: look at interaction plot
Ornstein2 <- Ornstein %>% group_by(sector, canada) %>%
  summarize(meanInterlocks = mean(interlocks), .groups = "keep")

ggplot(Ornstein2, aes(canada, meanInterlocks, group = sector)) + 
  geom_point() + geom_path() # some differences in slope, might expect an interaction - use type III sums of squares

# graphical: look at boxplots for homogeneity of variance
ggplot(Ornstein, aes(sector, interlocks, col = canada)) + 
  geom_boxplot() # some major differences in the shape of the boxplots, probably want to not assume homogeneity of variance
```

## 3

Conduct a two-way ANOVA on these data using options informed by the previous step, and calculate effect sizes. What do you find?

```{r echo = TRUE, eval = TRUE}
library(ez) # use ezANOVA for correct type III SS
Ornstein <- Ornstein %>% mutate(ID = factor(1:nrow(Ornstein))) # create ID variable

# White's adjustment
mod1 <- ezANOVA(data = Ornstein, dv = interlocks, wid = ID,
                between = c(sector, canada), white.adjust = TRUE, 
                type = 3, return_aov = TRUE) 
mod1 # no significant interaction, but both main effects are significant

# probably unwise to use WLS since that requires aov, which doesn't give us an easy option for type III SS

# effect sizes
library(sjstats)
anova_stats(mod1$aov)
# relatively small effects for both factors, sector has a larger effect than canada
```

## 4

Based on your results in the previous part, calculate either simple main effects or ordinary main effects. What do you conclude?

```{r echo = TRUE, eval = TRUE}
# want to do ordinary main effects since the interaction is not significant

Ornstein %>% group_by(canada) %>% summarize(mean(interlocks))
# Canadian companies have more interlocks on average than non-Canadian companies (makes sense)

# for sector, use any post-hoc test for a one-way ANOVA (pairwise t-test with holm shown here)
pairwise.t.test(x = Ornstein$interlocks, g = Ornstein$sector,
                p.adjust.method = "holm")
# significant differences between FIN vs. AGR, MAN, MER, and MIN
#   also between TRN vs. AGR and MAN

Ornstein %>% group_by(sector) %>% summarize(mean(interlocks))

# FIN has more interlocks than most other, TRN also, to some extent
```

