---
title: "Lab 13: Regression Extensions"
author: "PSYC 6802"
date: "December 10, 2025"
date-format: long
format: 
  revealjs:
    theme: sky
    scrollable: true
    smaller: true
    embed-resources: false
    slide-number: true
    chalkboard: true
    highlight-style: arrow
    link-external-icon: true
    html-math-method: mathjax
execute: 
  echo: true
  freeze: auto
---


## Packages for Today (Install if Needed)

```{r, message = FALSE}
library(tidyverse)
library(car)
library(lme4)
library(performance)
```

## Overview

Functions for logistic regression: 

* `glm(DV ~ IV1 + IV2 + ..., dat, family = binomial)`: fit the logistic regression model
* `coef(glm())` or `summary(glm())$coefficients`: extract coefficients for a glm model
* `fitted(glm())`: get fitted values (predicted probabilities)
* `predict(glm(), newdata = ..., type = "response")`: get predicted probabilities for new data scored with the fitted model
* `anova(glm1, test = "LRT")`: likelihood ratio test comparing `glm1` to the null model
* `anova(glm1, glm2, test = "LRT")`: likelihood ratio test comparing `glm1` to `glm2` (where one model is a subset of the other model) 

Function for multilevel regression:

* `lmer(DV ~ CV + (1 | ID), dat)`
  
    - `dat`: dataset in long format

    - `DV`: column name of dependent variable

    - `ID`: column name of user IDs or other cluster variable

    - `CV`: column name of covariate (may be IV)


## Data For Today

Today, we will try to predict volunteerism from personality characteristics. 

```{r}
data("Cowles", package = "carData")

Cowles %>% glimpse
```

## Exploring Data

Suppose we hypothesize that higher extraversion is associated with higher rates of volunteering.

We might first calculate summary statistics for each group (volunteer vs. not):

```{r}
Cowles %>% group_by(volunteer, sex) %>% 
  summarize(n(), 
            mean(neuroticism), 
            mean(extraversion), .groups = "keep")
```

From these, it seems as if people who volunteer have higher extraversion. 

## Logistic Regression

Let's next fit a logistic regression model with Extraversion as a predictor. 

The command to fit logistic regression is `glm()`. The formula notation is the same as for linear regression, but we need to add the argument `family = binomial` to tell R that we want logistic regression instead of another GLM.

```{r}
glm1 <- glm(volunteer ~ extraversion, Cowles, 
            family = binomial)
```

## Logistic Regression

```{r}
round(summary(glm1)$coefficients, 3)
```

These results imply the following fitted equation:

$\hat{P}(volunteer) = \frac{1}{1+\exp(-(-1.14 + .07Extraversion))}$

* -1.14 is the logit when Extraversion = 0 (highly introverted): $P = \frac{1}{1 + \exp(1.14)} = .24$

* .07 is the increase in logit (predicted log odds of volunteering) for a one-point increase in Extraversion score

## Coefficient interpretations

* .07 is the increase in logit (predicted log odds of volunteering) for a one-point increase in Extraversion score

Or

```{r}
exp(.07)
```

* For a one unit increase in Extraversion, volunteering becomes $exp(0.07) = 1.07$ times as likely. 

Sometimes $exp(0.07) = 1.07$ might also be written as $e^{0.07} = 1.07$

## Classification Accuracy

To evaluate accuracy, we need to get the fitted values for the training data.

```{r}
fitted1 <- fitted(glm1)
```

Then, we classify subjects as being predicted to volunteer or not based on a .5 cutoff.

```{r}
class1 <- ifelse(fitted1 > .5, "pred_yes", "pred_no")
table(class1, Cowles$volunteer)

(765 + 71)/(765 + 526 + 59 + 71)
```
We get around 59% accuracy which is not great. Extraversion may not be very informative in relation to the outcome. 

## Model Comparison


To test whether extraversion has "predictive power" associated with it, we conduct a likelihood ratio test comparing `glm1` to the null model (a model with only an intercept).

```{r}
anova(glm1, test = "LRT")

```

Here, we reject the null hypothesis that the null model fits as well as the model with extraversion. Therefore, we retain the model with extraversion as a predictor: $\chi^2(dif, 1) = 22.022, p < .001$.


## More than One Predictor

Perhaps we can construct a better model by including neuroticism as a predictor.

```{r}
glm2 <- glm(volunteer ~ extraversion + neuroticism, 
            Cowles, family = binomial)
round(summary(glm2)$coefficients, 3)
```

## Model Comparison

To conduct the likelihood ratio test, we can use the `anova()` function with the two models that we want to compare.

```{r}
anova(glm1, glm2, test = "LRT")
```

$\chi^2(dif, 1) = 0.936, p = .33$

Here, we *retain* the null hypothesis, meaning that the simpler model (the model without neuroticism) fits just as well as the more complex model.


## Wide vs. Long Data

Long data has as many rows, as few columns as possible

* Multiple rows for one subject
* One column for the DV
* One column for each IV

Wide data has fewer rows, more columns

* e.g., One row per subject
* Multiple columns for the DV at different IV levels

Long data is usually preferred for data analysis

* Some exceptions, such as for certain types of plots

## Dataset 1: Stroop Effect

We'll begin by working through the Stroop Effect data from lecture. Remember that this dataset gives the proportion of correct responses for three different levels of contingency between words and colors.

```{r}
library(readxl)
dat <- read_xlsx("Lab_13_Schmidt data.xlsx",
                 sheet = 2, range = "B20:E257", 
                 col_names = c("Subject", "High", "Medium", "Low"))
dat %>% glimpse
```

## Wide vs. Long Data

In this class, we have already seen some examples of moving between wide and long data.

tidyverse functions:

* `pivot_longer`: move from wide format to long format
* `pivot_wider`: move from long format to wide format

The Stroop data are stored in wide format. To change to long format:

```{r}
dat_long <- dat %>% pivot_longer(cols = -Subject,
                                 names_to = "Contingency", 
                                 values_to = "PropError")
```

The `cols = -Subject` argument means to exclude the `Subject` column in the information that is shifted to the new columns. Instead, the Subject variable will be stretched in the long format. See next slide for the structure of `dat_long`.

## Wide vs. Long Data

```{r}
dat_long %>% glimpse
```

* `Contingency` is the name of a new column taking values corresponding to the column names in the wide data set
* `PropError` is the name of a new column taking (usually numeric) values of the actual cell entries
* `Subject` is a variable storing information other than that stored in `Contingency` and `PropError`

## Wide vs. Long Data

To return to wide format:

```{r}
dat_long %>% pivot_wider(id_cols = Subject, 
                         names_from = Contingency,
                         values_from = PropError) %>% 
  glimpse
```
* `Contingency` name of (factor) variable in long data that contains wide format column names

* `PropError` name of variable in long data that contains cell values

## Data Cleaning

As a final piece of data cleaning, let's change the `Contingency` and `Subject` columns in `dat_long` to factors.

```{r}
dat_long <- dat_long %>% 
  mutate(Contingency = factor(Contingency),
         Subject = factor(Subject))

dat_long %>% glimpse
```


## Fitting multilevel models with `lmer`

To fit linear random-effects models, we can use the `lmer` function in the `lme4` package.

```{r}
mod <- lmer(PropError ~ (1 | Subject), dat_long)
```

This syntax means that `PropError` is the DV, `Subject` is the random effect (the mean for each subject is a random variable), and `dat_long` is the dataset. 

For now, this model doesn't account for the effect of `Contingency` level . It only models individual differences in the proportion of errors made.


## Fitting multilevel models with `lmer`

```{r echo = FALSE}
summary(mod)
```

* $\hat{\psi} = .013775$: estimated variance of the subject means (intercepts)

* $\hat{\theta} = .001966$: estimated residual (within-subject) variance

* $\hat{\beta}_0 = .344523$: estimated overall proportion of errors

## Intraclass Correlation Coefficient

```{r}
print(mod, ranef.comp = "Variance") # another way to view output
```

The ICC is the ratio of subject-level variance to total (subject + residual) variance: $\frac{\hat{\psi}}{\hat{\psi} + \hat{\theta}}$

```{r}
0.013775 / (0.013775 + 0.001966)
```

87.5% is the proportion of variance associated with individual differences in error rates

## Intraclass Correlation Coefficient

The `performance` package also has an `icc()` function that can be called on `lmer` models. It will give you two values. We will generally want the **adjusted** ICC, though when there are no predictors they will be the same

```{r}
icc(mod)
```

## REML vs. ML

`lmer` uses REML estimation by default. Set `REML = FALSE` for ML estimation.

```{r}
summary(lmer(PropError ~ (1 | Subject), dat_long, REML = FALSE))
```

Some differences in random-effects estimates, but REML is preferred, especially when the number of clusters (persons) is small.

## HLM With a Level-1 Covariate

Next, let's add a fixed effect of `Contingency`. Similar to other regression coefficients, this assumes a linear relationship between `Contingency` and `PropError`. 

```{r}
mod2 <- lmer(PropError ~ Contingency + (1 | Subject), dat_long)
print(mod2, ranef.comp = "Variance")
```

::: columns
:::: column

* $\hat{\psi} = 0.013831$: estimated variance of the subject means (intercepts)

* $\hat{\theta} = 0.001796$: estimated residual (within-subject) variance

::::
:::: column

* $\hat{\beta}_0 = 0.33118$: estimated error proportion at high contingency

* $\hat{\beta}_1 = 0.02659$: estimated increase in error rate for low vs. high contingency

* $\hat{\beta}_2 = 0.01345$: estimated increase in error rate for medium vs. high contingency

::::
:::

## Residual Intraclass Correlation

* $\hat{\psi} = 0.013831$: estimated variance of the subject means (intercepts)

* $\hat{\theta} = 0.001796$: estimated residual (within-subject) variance

```{r}
icc(mod2) # we want adjusted ICC here
```

After accounting for the different trial types, 87.8% of the remaining variance is associated with individual mean differences.

## Likelihood Ratio Test to Compare Models

The `anova` function will compare two competing models (fit to the same data) via a number of measures, including the likelihood ratio test. 

```{r}
anova(mod, mod2)
```

These results show significantly better fit for the model that includes `Contingency` as a covariate. There are 2 degrees of freedom for this test, associated with the 2 added dummy variables associated with the 3-category covariate in `mod2`.

$\Delta \chi^2 (2) = 44.85, p < .001$
